PMV: Comparing the prediction accuracy of ISO 7730:2005 and ASHRAE 55:2023

### 2023-09-12 Manuscript/src/sections/conclusions.tex
feat: finalised second draft to send to Stefano

\section{Conclusions}\label{sec:conclusions}
The present study aimed to compare the accuracy of the \ac{pmv} model, as used in the \gls{7730} Standard, and its modified version, the \gls{pmv-ce} model, as employed in the \gls{55} Standard.
Our findings revealed notable disparities in the predictive accuracy of the \ac{pmv} and \gls{pmv-ce} models when compared against the \ac{tsv} available in the ASHRAE Global Thermal Comfort Database II v2.1 (Comfort DB).
These disparities shed light on the limitations and strengths of the \ac{pmv} model, enabling a more informed decision-making process for the selection of an appropriate formulation within National and International Standards.
The outcomes of this analysis have significant implications for researchers, practitioners, and policymakers involved in the field of thermal comfort assessment and standardization.

>>>>

\section{Conclusions}\label{sec:conclusions}
The present study compares the accuracy of the \ac{pmv} model, as used in the \gls{7730} Standard, and its modified version, the \ac{pmv-ce} model, as employed in the \gls{55} Standard.
The \ac{pmv-ce} was specifically developed with the aim of more accurately estimating latent and sensible heat losses from the skin to the environment when air movement is used~\cite{arens_moving_2009}.
We determined the predictive accuracy of the \ac{pmv} and \ac{pmv-ce} models by comparing their results to the \ac{tsv} available in the ASHRAE Global Thermal Comfort Database II v2.1 (Comfort DB).

The overall accuracy of the \ac{pmv} and \ac{pmv-ce} were \qty{33}{\percent} and \qty{34}{\percent}, respectively.
Both models had an accuracy lower than \qty{2}{\percent} when predicting either `warm', `hot', or `cold'.
This value is significantly lower and worse than random guessing, this is very concerning because this means that the models are actively misleading the users.
When only comparing the results of the two models for entries with \ac{v}~$\geq$~\qty{0.2}{\m\per\s} -- range in which the two models differ -- the \ac{pmv} outperformed the \ac{pmv-ce} on different metrics.
For example, the median bias of the \ac{pmv} model is \var{bias_median_pmv_0.2} while the median bias for the \ac{pmv-ce} is \var{bias_median_pmv_ce_0.2}.

Our results also show that the classic assumption that people who are `slightly warm' or `slightly cool' are thermally comfortable leads to under-counting thermal discomfort in our dataset.
In the \ac{db2} \qty{68}{\percent} of participants who were `slightly warm' wanted to be `cooler' and \qty{54}{\percent} of participants who were `slightly cool' wanted to be `warmer'.
This finding challenges the above assumption.

\subsection{Implications for the ASHRAE 55 and ISO 7730 Standards}\label{subsec:implications-for-the-ashrae-55-and-iso-7730-standards}
Choosing between the \ac{pmv} and \ac{pmv-ce} is a source of confusion wordwide since both models are widely adopted and incorporated in many building codes and Standards worldwide.
The \gls{55} Standard explains how the \ac{pmv-ce} works, however, we could not find a peer-reviewed publication that clearly explain the benefit of using it.
The \ac{pmv-ce} is more complex, computationally intensive to run, while not being more accurate than the original \ac{pmv} model.
Hence we do not recommend its use.

The low prediction accuracy of both model formulations, coupled with the lack of data in the extremes of the applicability limits of the models, may suggest that the models' applicability limits should be revisited.
Based on our results we recommend restricting the use of the \ac{pmv} only to a values $\lvert \textrm{PMV}\lvert \leq 0.5$.
Or in other words, we only recommend its use to determine if an environment may be deemed comfortable by a large group of occupants.
The model, however, is not capable of determining the degree by which people are dissatisfied and can only be used to classy if people are on the `cold' (\ac{pmv}~$\leq 0.5$) or `warm' (\ac{pmv}~$\geq 0.5$) side.
Several peer-reviewed papers \cite{Cheung2019, Yao2022, kim2019thermal, tartarini2018thermal, Humphreys2002, doherty_evaluation_1988, tartarini_prediction_2023} provide corroborating evidence that support our results.

We also recommend restricting the ranges on both environmental and personal factors set by both \gls{55} and \gls{7730} Standards.
Currently, these limits extend far beyond the range of data contained in the \ac{db2} dataset.
Being the \ac{db2} the largest thermal comfort dataset in the world and containing data from all continents we deem not to be necessary to have such wide applicability limits for both \gls{55} and \gls{7730} Standards when these environmental variables are rarely observed in buildings across the world.
In addition, the lack of data available beyond the ranges depicted in Figure~\ref{fig:dist_input_data} did not allow us to test the accuracy of both the \ac{pmv} and \ac{pmv-ce} above those ranges.
Consequently, we suggest reducing both Standards applicability limits to the ranges shown in Table~\ref{tab:ranges}, until more data are collected.
\begin{table}[htb!]
    \centering
    \begin{tabular}{cc}
        \toprule
        Variable & Proposed range \\
        \midrule
        \ac{tdb} & \qtyrange{17}{30}{\celsius} \\
        \ac{tr} & \qtyrange{17}{30}{\celsius} \\
        \ac{rh} & \qtyrange{20}{80}{\percent} \\
        \ac{clo} & \qtyrange{0.3}{1.5}{clo} \\
        \ac{met} & \qtyrange{1}{2}{met} \\
        % todo shall we also include limits on Age?
        % todo shall we also include airspeed?
        \bottomrule
    \end{tabular}
    \caption{New proposed applicability limits for the \gls{55} and \gls{7730} Standards.}
    \label{tab:ranges}
\end{table}
\mycite{Humphreys2002} reached a similar conclusion and observed that the ranges of inputs that are consistent with a valid use of PMV are much narrower than those given in \gls{7730}.
The limit on \ac{met} is also corroborated by evidence provided by \mycite{doherty_evaluation_1988}.

The \ac{tsv} is ambiguous and does not allow determining if an action should be taken to increase the participant's thermal comfort.
Assuming that people who are `slightly warm' or `slightly cool' are comfortable leads to under-counting thermal discomfort in our dataset, and it is not a valid assumption.
We recommend the use of the thermal preference scale to determine how people perceive their thermal environment, this scale more clearly allows an HVAC system to decide whether to change or not the thermal environment.

### 2023-09-12 Manuscript/src/sections/introduction.tex
feat: finalised second draft to send to Stefano

To this date, the \ac{pmv} still remains to be the most utilized thermal comfort model even though several studies have highlighted that the \ac{pmv} has low accuracy in predicting thermal sensation votes~\cite{Cheung2019, Yao2022, kim2019thermal, tartarini2018thermal, Humphreys2002, doherty_evaluation_1988}.

To overcome the accuracy concerns, in particular when high air speed is used to cool people, the \gls{55} Standard uses a modified version of the original model, the \ac{pmv-ce}~\cite{ashrae552023}.
The best source to see how \ac{pmv-ce} works is the Standard itself, but the standard does not explain why some choices were implemented.
A partial justification of the model is described in \mycite{arens_moving_2009} and \mycite{yang_cooling_2015}.
The process used to calculate the \ac{pmv-ce} in the \gls{55} is summarized in the flowchart depicted in Figure~\ref{fig:flowchart_pmv_ce}.
In summary, when the \ac{v} exceeds \qty{0.1}{\m\per\s} the \gls{55} prescribes the use of the \ac{ce}.
The value of \ac{ce} is calculated using the \ac{set} equation, and it is then subtracted from both the \ac{tdb} and \ac{tr}.
The resulting values become the new inputs in the \ac{pmv} model.
Since the \ac{ce} accounts for the convective heat losses from the person to its environment the value of \ac{v}, used to calculate the \ac{pmv}, is set to \qty{0.1}{\m\per\s}.
The other three input parameters (\ac{clo}, \ac{met}, and \ac{rh}) remain unchanged.
Consequently, for a given thermal environment, results of the two \ac{pmv} formulations differ when the value of \ac{v} is higher than \qty{0.1}{\m\per\s}.
%The rationale for this approach is that the original formulation of the \ac{pmv} model cannot accurately estimate latent and sensible heat losses from the skin to the environment when \ac{v} is higher than \qty{0.1}{\m\per\s}.

@@

        \draw [arrow] (start) -- (dec1);
        \draw [arrow] (dec1) -- node[above,pos=0.3] {Yes} (pro1a);
        \draw [arrow] (dec1) -- node[right,pos=0.3] {No} (dec2);
        \draw [arrow] (pro1a) -- ($(pro1a.south)+(0,-0.5)$) -| (dec2);
        \draw [arrow] (dec2) -- node[above,pos=0.3] {Yes} (pro2a);
        \draw [arrow] (dec2) -- node[right,pos=0.3] {No} (dec3);
        \draw [arrow] (pro2a) -- ($(pro2a.south)+(0,-0.5)$) -| (dec3);
        \draw [arrow] (dec3) -- ($(dec3.west)$) -| node[above,pos=0.3] {No} (end1);
        \draw [arrow] (dec3) -- ($(dec3.east)$) -| node[above,pos=0.3] {Yes} (pro3);
        \draw [arrow] (pro3) -- (end2);
    \end{tikzpicture}
    \caption{Flowchart depicting the steps for the calculation of the PMV in accordance with the \acs{55} Standard.}
    \label{fig:flowchart_pmv_ce}
\end{figure}

@@

Some notable works that tried to determine the accuracy of the \ac{pmv} model include the manuscript from \mycite{doherty_evaluation_1988} who evaluated the ability of the \ac{pmv} and \ac{set} models to predict several physiological variables (i.e., skin temperature, core temperature, and skin wettedness) under a wide range of still air environments and metabolic rates.
They concluded that the \ac{pmv} model is accurate for simulations of resting subjects, but its accuracy decreases as a function of metabolic rate.
% todo use this finding to support limit to PMV range
Humphreys and Nicol estimated the effects of measurement and formulation error on predicting thermal sensation using the \ac{pmv}~\cite{Humphreys2000}.
They used the ASHRAE Global Thermal Comfort database I and determined that the measurement error and the error introduced by the \ac{pmv} model formulation had a similar and non-negligible contribution.
They also determined the validity of the \ac{pmv} for predicting comfort votes collected in field studies~\cite{Humphreys2002}.
They concluded that the \ac{pmv} range of applicability should be significantly reduced and it fails to predict the extent of thermal dissatisfaction of people in buildings.
% todo use this finding to support limit to PMV range
The \ac{pmv} was free from bias only when it was used to predict thermal neutrality.
\mycite{Cheung2019} determined the accuracy of the \ac{pmv} model, by comparing its results with the \ac{tsv} from the ASHRAE Global Thermal Comfort Database II.
They found that the thermal sensation predicted by the PMV model, on average, is one full thermal sensation scale unit away from the subject’s responses, confirming the results of~\cite{Humphreys2002}.
The accuracy of PMV was only \qty{34}{\percent}, the model has a slightly higher prediction accuracy for sensation close to neutrality, but the accuracy declined towards either end of the thermal sensation scale, and it overestimated both hot and cold sensations.
They also found that the Predicted Percentage of Dissatisfied (PPD) failed to predict the percentage of unacceptable votes if the thermal sensation was predicted using the PMV model and suggested its removal from \gls{55} Standards.
These results were confirmed based on the Chinese Thermal Comfort database~\cite{du_evaluation_2022}.
\mycite{Yao2022} compared the \ac{pmv} and \ac{pmv-ce} models, however, their aim was primarily to compare these two formulations with other adaptive \ac{pmv} formulations, hence, they do not provide a detailed analysis on the prediction accuracy of the two models.
They focus significantly on how the models perform in different climates and when applied to people from different world regions and their analysis only reports the mean bias of the different models.
This, as depicted by \mycite{Humphreys2000}, does not provide sufficient insights and information in determining which model is more accurate since it does not explain how the model performs over a wide range of environmental, personal conditions, and \ac{tsv}.
Reporting the classification accuracy of the \ac{pmv} formulations when people are grouped by their \ac{tsv} is particularly for an unbalanced dataset like the \ac{db2} where most of the participants reported to be `neutral'.
Finally, they only reported the overall bias for the whole dataset, despite the fact that the two formulations only differ when the \ac{v} exceeds \qty{0.1}{\m\per\s}.

>>>>

To this date, the \ac{pmv} still remains to be the most utilized thermal comfort model even though several studies have highlighted that the \ac{pmv} has low accuracy in predicting thermal sensation votes~\cite{Cheung2019, Yao2022, kim2019thermal, tartarini2018thermal, Humphreys2002, doherty_evaluation_1988, tartarini_prediction_2023}.

To overcome the accuracy concerns, in particular when high air speed is used to cool people, the \gls{55} Standard uses a modified version of the original model, the \ac{pmv-ce}~\cite{ashrae552023}.
The best source to see how \ac{pmv-ce} works is the Standard itself, but the Standard does not explain why some choices were implemented.
A partial justification of the model is described in \mycite{arens_moving_2009} and \mycite{yang_cooling_2015}.
The process used to calculate the \ac{pmv-ce} in the \gls{55} is summarized in the flowchart depicted in Figure~\ref{fig:flowchart_pmv_ce}.
In summary, when the \ac{v} exceeds \qty{0.1}{\m\per\s} the \gls{55} prescribes the use of the \ac{ce}.
The value of \ac{ce} is calculated using the \ac{set} equation, and it is then subtracted from both the \ac{tdb} and \ac{tr}.
The resulting values become the new inputs in the \ac{pmv} model.
Since the \ac{ce} accounts for the convective heat losses from the person to its environment the value of \ac{v}, used to calculate the \ac{pmv}, is set to \qty{0.1}{\m\per\s}.
The other three input parameters (\ac{clo}, \ac{met}, and \ac{rh}) remain unchanged.
Consequently, for a given thermal environment, results of the two \ac{pmv} formulations differ when the value of \ac{v} is higher than \qty{0.1}{\m\per\s}.
%The rationale for this approach is that the original formulation of the \ac{pmv} model cannot accurately estimate latent and sensible heat losses from the skin to the environment when \ac{v} is higher than \qty{0.1}{\m\per\s}.

@@

        \draw [arrow] (start) -- (dec1);
        \draw [arrow] (dec1) -- node[above,pos=0.3] {Yes} (pro1a);
        \draw [arrow] (dec1) -- node[right,pos=0.3] {No} (dec2);
        \draw [arrow] (pro1a) -- ($(pro1a.south)+(0,-0.5)$) -| (dec2);
        \draw [arrow] (dec2) -- node[above,pos=0.3] {Yes} (pro2a);
        \draw [arrow] (dec2) -- node[right,pos=0.3] {No} (dec3);
        \draw [arrow] (pro2a) -- ($(pro2a.south)+(0,-0.5)$) -| (dec3);
        \draw [arrow] (dec3) -- ($(dec3.west)$) -| node[above,pos=0.3] {No} (end1);
        \draw [arrow] (dec3) -- ($(dec3.east)$) -| node[above,pos=0.3] {Yes} (pro3);
        \draw [arrow] (pro3) -- (end2);
    \end{tikzpicture}
    \caption{Flowchart depicting the steps for the calculation of the PMV in accordance with the \gls{55} Standard.}
    \label{fig:flowchart_pmv_ce}
\end{figure}

@@

Some notable works that tried to determine the accuracy of the \ac{pmv} model include the manuscript from \mycite{doherty_evaluation_1988} who evaluated the ability of the \ac{pmv} and \ac{set} models to predict several physiological variables (i.e., skin temperature, core temperature, and skin wettedness) under a wide range of still air environments and metabolic rates.
They concluded that the \ac{pmv} model is accurate for simulations of resting subjects, but its accuracy decreases as a function of metabolic rate.
Humphreys and Nicol estimated the effects of measurement and formulation error on predicting thermal sensation using the \ac{pmv}~\cite{Humphreys2000}.
They used the ASHRAE Global Thermal Comfort database I and determined that the measurement error and the error introduced by the \ac{pmv} model formulation had a similar and non-negligible contribution.
They also determined the validity of the \ac{pmv} for predicting comfort votes collected in field studies~\cite{Humphreys2002}.
They concluded that the \ac{pmv} range of applicability should be significantly reduced and it fails to predict the extent of thermal dissatisfaction of people in buildings.
The \ac{pmv} was free from bias only when it was used to predict thermal neutrality.
\mycite{Cheung2019} determined the accuracy of the \ac{pmv} model, by comparing its results with the \ac{tsv} from the ASHRAE Global Thermal Comfort Database II.
They found that the thermal sensation predicted by the PMV model, on average, is one full thermal sensation scale unit away from the subject’s responses, confirming the results of~\cite{Humphreys2002}.
The accuracy of PMV was only \qty{34}{\percent}, the model has a slightly higher prediction accuracy for sensation close to neutrality, but the accuracy declined towards either end of the thermal sensation scale, and it overestimated both hot and cold sensations.
They also found that the Predicted Percentage of Dissatisfied (PPD) failed to predict the percentage of unacceptable votes if the thermal sensation was predicted using the PMV model and suggested its removal from \gls{55} Standards.
These results were confirmed based on the Chinese Thermal Comfort database~\cite{du_evaluation_2022}.
\mycite{Yao2022} compared the \ac{pmv} and \ac{pmv-ce} models, however, their aim was primarily to compare these two formulations with other adaptive \ac{pmv} formulations, hence, they do not provide a detailed analysis on the prediction accuracy of the two models.
They focus significantly on how the models perform in different climates and when applied to people from different world regions and their analysis only reports the mean bias of the different models.
This, as depicted by \mycite{Humphreys2000}, does not provide sufficient insights and information in determining which model is more accurate since it does not explain how the model performs over a wide range of environmental, personal conditions, and \ac{tsv}.
Reporting the classification accuracy of the \ac{pmv} formulations when people are grouped by their \ac{tsv} is particularly for an unbalanced dataset like the \ac{db2} where most of the participants reported to be `neutral'.
Finally, they only reported the overall bias for the whole dataset, despite the fact that the two formulations only differ when the \ac{v} exceeds \qty{0.1}{\m\per\s}.
\mycite{tartarini_prediction_2023} also published a conference paper on this topic.
They concluded that the \ac{pmv} and \ac{pmv-ce} models accuracy is \qty{35}{\percent} and \qty{36}{\percent} for $\mid$\ac{pmv}$\mid \leq 2$, \qty{43}{\percent} and \qty{44}{\percent} for $\mid$\ac{pmv}$\mid \leq 1$ in predicting the self-reported thermal sensation of people, respectively.
However, their analysis is limited to providing the precision of the model alone, without reporting the sensitivity (also commonly referred as recall) and they do not provide the bias of the two models.
